---
title: "agricultural_data"
author: "owen flanagan"
date: "30 April 2019"
output: 
  github_document:
    toc: true
    toc_depth: 2

---

# Introduction

This work is a study of some interesting agricultural indicators I came across while working on some analysis of economic freedom. 
I believe that explaining techniques, their use, and justifying them is an important part of any data scientists workflow so I will aim to explain my reasoning and justification - practice makes perfect.

The data I will be using in this post were sourced from the World Bank. A full list of indicators can be found here:
https://data.worldbank.org/indicator

Of these indicators I picked out a subset which I thought would be interesting together,

https://data.worldbank.org/indicator/AG.LND.IRIG.AG.ZS?view=chart
Agricultural irrigated land % of total agricultural land

https://data.worldbank.org/indicator/AG.LND.TRAC.ZS?view=chart
Agricultural machinery, tractors per 100 sq. km of arable land
more economic freedom, more machines
*** "Quote some of the invisible hand here" ***

https://data.worldbank.org/indicator/AG.YLD.CREL.KG?view=chart
Cereal yield (kg per hectare)
interesting to look at wrt tractora and irrigation and fertilizer
might be interesting to build some models predicting cereal yield as target - glm and rf with feature importance

https://data.worldbank.org/indicator/AG.CON.FERT.ZS?view=chart
Fertilizer consumption (kilograms per hectare of arable land)

Taken together we have a dataset which contains:
irrigation 
fertilizer
mechanisation
and resultant cereal yield

This seems like a reasonable application of modelling and I will apply both linear models and tree ensembles to this data set to see what level of predictive accuracy I can get. 

(the countries dataset from world bank might include some other interesting data such as wealth, lat and longitude and similar)

# Data loading and tidying
The data we are planning on using is provided by the world bank. The world bank does provide an API which we could query, but there is an R package available called wbstats which provides easy access to the data we want.
Information on using that wbstats package can be found here:

https://cran.r-project.org/web/packages/wbstats/vignettes/Using_the_wbstats_package.html

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#import our libraries
library(tidyverse) # tidyverse provides a beautiful set of tools for data science with consistent syntax and behaviour
library(wbstats) #wbstats provides access to the world bank data sets
library(knitr) #provides a set of tools to customize the processing of Rmarkdown including kable - a very simple table generator
library(kableExtra) # provides a set of tools to extend kable -
```

## Loading


```{r}
countries <- wb_cachelist$countries %>% as.tbl()
countries <- countries %>% 
  select(-c(capital,iso2c,region_iso2c,regionID,adminID,admin_iso2c,incomeID,income_iso2c,lendingID,lending_iso2c,lending,admin)) %>% 
  mutate(long = as.numeric(long), lat = as.numeric(lat))
countries %>% 
  head() %>% 
  kable()
```


```{r}
irrigation <- wb(indicator = "AG.LND.IRIG.AG.ZS") %>% #download the specified indicator from the World Bank API
  as.tbl() %>% #convert the dataframe to a tibble for better console printing in interactive analysis (limits default number of rows printed to a reasonable number and prints var type of each column beneath row name)
  mutate(date = as.numeric(date)) #convert the date column to a numeric value

irrigation %>% 
  head()%>% #return the first 6 rows of the irrigation tibble as a new tibble
  kable() #output the new tibble as an html friendly table
```



```{r}
tractors <- wb(indicator = "AG.LND.TRAC.ZS") %>% 
  as.tbl() %>% 
  mutate(date = as.numeric(date))
tractors %>% head() %>% kable()
```


```{r}
fertilizer <- wb(indicator = "AG.CON.FERT.ZS") %>% 
  as.tbl()%>% 
  mutate(date = as.numeric(date))
fertilizer  %>% head() %>% kable()
```


```{r}
cereal_yield <- wb(indicator = "AG.YLD.CREL.KG") %>% 
  as.tbl() %>% 
  mutate(date = as.numeric(date)) 
cereal_yield %>% head() %>% kable()
```


Some notes on what we can see so far:

 * We essentially have a number of data tables with the format [Country,Year,Indicator,Value] with some of these repeated in different formats. Needs reshaping.
 * We don't really have a very good name column at the moment. Need to make.
 * The country column for several of these datasets has a value "Arab World". As far as I am aware, that's not a country. Needs cleaning.

 
## Reshaping
 
```{r}
irrigation_red <- irrigation %>% 
  select(iso3c, country, date, value) %>% 
  rename(IrrigationPercentOfLand = value)

irrigation_red %>% head() %>% kable()
```

```{r}
tractors_red <- tractors %>% 
  select(iso3c, country, date, value) %>% 
  rename(TractorsPer100Sqkm = value)

tractors_red %>% head() %>% kable()
```

```{r}
fertilizer_red <- fertilizer %>% 
  select(iso3c, country, date, value) %>% 
  rename(FertilizerKgPerHectare = value)
fertilizer_red %>% head() %>% kable()
```

```{r}
cereal_yield_red <- cereal_yield %>% 
  select(iso3c, country, date, value) %>% 
  rename(CerealYieldKgPerHectare = value)
cereal_yield_red %>% head() %>% kable()
```

## Removing non countries

The easiest way to remove the non-countries from the data set is to join onto the countries dataframe.









```{r}
df <- countries %>%
    left_join(irrigation_red %>% select(-country),by="iso3c") %>% 
    left_join(tractors_red %>% select(-country),by=c("iso3c","date")) %>% 
    left_join(fertilizer_red %>% select(-country),by=c("iso3c","date")) %>% 
    left_join(cereal_yield_red %>% select(-country),by=c("iso3c","date"))

```



```{r}
df %>% head() %>% kable()
```

```{r}
df %>% summary()
```

We can see we have many NAs in our dataset. It seems that a lot of rows are missing longitude and lattitude data, and more are missing date. My guess would be that a lot of rows are missing values in a lot of columns and these are for countries like Aruba. We should start by looking at the columns where the data is missing from the fewest rows. This will allow us to potentially reduce the Na's from most rows.

```{r}
df %>% filter(is.na(long)) %>% head() %>%  kable()

```

When we do this we see that these are indeed rows for non-countries - we don't care about them.

```{r}
df <- df %>% filter(!is.na(long))
df %>% summary()
```

Working off the same strategy of investigating the columns with the least rows missing we can look at date.

```{r}
df %>% filter(is.na(date)) %>% head() %>%  kable()
```

These rows are for countries that we may actually be interested in. It may be that there is simply not data available for these countries, or it may be something else. This deserves investigation.

```{r}
countries_with_one_row <- irrigation_red %>% 
  #group dataframe by country name
  group_by(country) %>% 
  #count the number of rows - this is the number of years of data
  summarise(n=n()) %>% 
  #filter to only have countries with a single row
  filter(n == 1)

irrigation_red %>% 
  #filter the original data set to only contain the rows for countries with single rows
  filter(country %in% countries_with_one_row$country) %>%
  head() %>% 
  kable()

```

When we do this analysis, we see that the issue is that the dataset does not have data for all years for all countries.
We can get a better idea of this by making a histogram.

```{r}
irrigation_red %>% 
  #group dataframe by country name
  group_by(country) %>% 
  #count the number of rows - this is the number of years of data
  summarise(n=n()) %>% 
  ggplot(aes(n)) + geom_histogram() + 
  ggtitle("Number of years data per country in irrigation dataset")

```

We can see straight away that there are only 13 countries in our data set which span the whole range.
At this stage we could ask if the countries with less than the full range are a continuous range (i.e. 2005-2008), or a number of sparse points accross the 16 years.
Before digging deeper into that minor question, it would be better to investigate this for the other data sets.

```{r}
tractors_red %>% 
  #group dataframe by country name
  group_by(country) %>% 
  #count the number of rows - this is the number of years of data
  summarise(n=n()) %>% 
  ggplot(aes(n)) + geom_histogram() +
  ggtitle("Number of years data per country in tractors dataset")

```

```{r}
fertilizer_red %>% 
  #group dataframe by country name
  group_by(country) %>% 
  #count the number of rows - this is the number of years of data
  summarise(n=n()) %>% 
  ggplot(aes(n)) + geom_histogram() + 
  ggtitle("Number of years data per country in fertilizer dataset")
```

```{r}
cereal_yield_red %>% 
  #group dataframe by country name
  group_by(country) %>% 
  #count the number of rows - this is the number of years of data
  summarise(n=n()) %>% 
  ggplot(aes(n)) + geom_histogram() +
  ggtitle("Number of years data per country in cereal yield dataset")
```

It is interesting that cereal_yield and tractors have a max number of rows per country of approximately 50 while fertilzier and irrigation ahve approx 16.

```{r}
cereal_yield_red %>% 
  group_by(country) %>% 
  summarise(n=n()) %>% 
  arrange(desc(n))
```

Afghanistan has a 57 rows so we might as well pick that.
An easy way of getting an idea of the full range of the data is a quick plot.

```{r}
cereal_yield_red %>% filter(country=="Afghanistan") %>% 
  ggplot(aes(x=date,y=CerealYieldKgPerHectare)) + 
  geom_point() + 
  geom_line() +
  ggtitle("Cereal Yield in Afghanistan overlaid with key events")+
  #Monarchy deposed
  geom_vline(xintercept =  1973,color="sky blue") +
  #start of Soviet Afghan War
  geom_vline(xintercept =  1979,color="red") +
  #end of Soviet Afghan War
  geom_vline(xintercept = 1989,color="red") +
  #Start of civil war
  geom_vline(xintercept = 1992,color="dark green") +
  #end of civil war and start of Taliban rule
  geom_vline(xintercept = 1996,color="dark green") +
  #Start of Operation Enduring Freedom
  geom_vline(xintercept = 2001,color="blue") 
  
```

This gives us what is quite an interesting graph for a region that has suffered decades of turmoil. This provides us with a story that can provide some motivation for this work. What is driving these changes in cereal yields?

*** NOTE - SHOULD PUT THIS GRAPH AT THE START TO TRY AND PROVIDE A SOURCE OF MOTIVATION FOR THE STORY ***

A quick bit of research on wikipedia allowed me to add lines roughly seperating the different periods of recent Afghan history.

 * Afghanistan was ruled by monarchy until the last king Mohammed Zahir Shah was overthrown by his cousin Mohammed Daoud Khan
in 1973.

 * In 1978 Afghanistan, the Saur Revolution led to an unpopular, aggresively modernising, socialist government backed by the USSR military.

 * By 1989 the USSR had given up on its campaign in the region.
From 1992-1996 there was a civil war which ended with the Taliban taking control.

 * In 2001 the USA and it's allies launched Operation Enduring Freedom as a response to the September 11 terrorist attacks and the beginning of a massive campaign of rebuilding and investing in the region.

As we have the other data sets we can have a quick look at the fertilizer yield for Afghanistan.

```{r}
fertilizer_red %>% 
  filter(country=="Afghanistan") %>% 
  ggplot(aes(x=date,y=FertilizerKgPerHectare)) + 
  geom_point() + 
  geom_line() +
  ggtitle("FertilizerPerHectare in Afghanistan")
```

```{r}
tractors_red %>% filter(country == "Afghanistan") %>% 
  ggplot(aes(x=date,y=TractorsPer100Sqkm)) +
  geom_point()+
  geom_line()+
  ggtitle("Tractor usage in Afghanistan")
```
```{r}
irrigation_red %>%
  filter(country=="Afghanistan") %>% 
  ggplot(aes(x=date,y=IrrigationPercentOfLand)) +
  geom_point() +
  geom_line()+
  ggtitle("Irrigation in Afghanistan")
```

As individual plots, these do not tell much of a story. We should combine them all together.

```{r}

afghan_cereal <- cereal_yield_red %>% 
  filter(country=="Afghanistan")
afghan_irrigation <- irrigation_red %>%
  filter(country=="Afghanistan")
afghan_fertilizer <- fertilizer_red %>% 
  filter(country=="Afghanistan")
afghan_tractors <- tractors_red %>% 
  filter(country=="Afghanistan")

afghan_combined <- afghan_cereal %>% 
  left_join(afghan_fertilizer) %>% 
  left_join(afghan_irrigation) %>% 
  left_join(afghan_tractors)

afghan_combined %>% 
  select(-c(iso3c,country)) %>% 
  gather(key,value,-date) %>% 
  ggplot(aes(x=date,y=value)) +
  facet_wrap(facets=vars(key),ncol=1, scales="free_y")+
  geom_point() +
  geom_line()

```

When taken together we can see that these datasets do not have a significant period of overlap for Afghanistan.  


## Returning to the histograms

```{r, echo=F}
irrigation_red %>% 
  #group dataframe by country name
  group_by(country) %>% 
  #count the number of rows - this is the number of years of data
  summarise(n=n()) %>% 
  ggplot(aes(n)) + geom_histogram() + 
  ggtitle("Number of years data per country in irrigation dataset")

```

```{r, echo=F}
tractors_red %>% 
  #group dataframe by country name
  group_by(country) %>% 
  #count the number of rows - this is the number of years of data
  summarise(n=n()) %>% 
  ggplot(aes(n)) + geom_histogram() +
  ggtitle("Number of years data per country in tractors dataset")

```

```{r, echo=F}
fertilizer_red %>% 
  #group dataframe by country name
  group_by(country) %>% 
  #count the number of rows - this is the number of years of data
  summarise(n=n()) %>% 
  ggplot(aes(n)) + geom_histogram() + 
  ggtitle("Number of years data per country in fertilizer dataset")
```

```{r, echo=F}
cereal_yield_red %>% 
  #group dataframe by country name
  group_by(country) %>% 
  #count the number of rows - this is the number of years of data
  summarise(n=n()) %>% 
  ggplot(aes(n)) + geom_histogram() +
  ggtitle("Number of years data per country in cereal yield dataset")
```

When we examine each of these histograms in turn we see that the cereal yield and fertilizer datasets mostly consist of countries with data points for all countries in the years they cover. 
The fertilizer and tractor datasets cover a far shorter range of time than the cereal yield - approx 15 years vs 50 years. 
If we want to study the relationship of these variables we are only going to be able to do it for the 15 years where both data sets exist.
This same logic applies for our other datasets (tractors and irrigation).


# next steps
 * ensure all data is joined together correctly and a nice table is selected
 * perform some modelling in a new file (part 2). Use tidymodel tools,
 start with linear models and test assumptions - look at residuals, find outliers,
 find good ways to plot residuals and outliers, look at distributions,
 * look at tools from the StatisticalModels and replicated some of the techniques
 * after that make a new file (part 3) with tree based methods. start with some decisions trees and investigate them. 
 * Then move to random forest models as per Jeremy Howard
 * (part 4) also use the feature importance techniques with linear models. At this point can potentially also throw some other model types at it, demonstrate parsnip. Should include the (decision tree + linear model complex model) and look into what packages exist around it - if there is no package for random forest then write one in parralel.



cereal
