---
title: "agricultural_data"
author: "owen flanagan"
date: "30 April 2019"
output: 
  github_document:
    toc: true
    toc_depth: 2

---

#Introduction
This work is a study of some interesting agricultural indicators I came accross while working on some analysis of economic freedom. 
I believe that explaining techniques, their use, and justifying them is an important part of any data scientists workflow so I will aim to explain my reasoning and justification - practice makes perfect.

The data I will be using in this post were sourced from the World Bank. A full list of indicators can be found here:
https://data.worldbank.org/indicator

Of these indicators I picked out a subset which I thought would be interesting together,

https://data.worldbank.org/indicator/AG.LND.IRIG.AG.ZS?view=chart
Agricultural irrigated land % of total agricultural land

https://data.worldbank.org/indicator/AG.LND.TRAC.ZS?view=chart
Agricultural machienry, tractors per 100 sq. km of arable land
more economic freedom, more machines
"Quote some of the invisible hand here"

https://data.worldbank.org/indicator/AG.YLD.CREL.KG?view=chart
Cereal yield (kg per hectare)
interesting to look at wrt tractora and irrigation and fertilizer
might be interesting to build some models predicting cereal yield as target - glm and rf with feature importance

https://data.worldbank.org/indicator/AG.CON.FERT.ZS?view=chart
Fertilizer consumption (kilograms per hectare of arable land)

Taken together we have a dataset which contains:
irrigation 
fertilizer
mechanisation
and resultant cereal yield

This seems like a reasonable application of modelling and I will apply both linear models and tree ensembles to this data set to see what level of predictive accuracy I can get. 

(the countries dataset from world bank might include some other interesting data such as wealth, lat and longitude and similar)

#Data loading and tidying
The data we are planning on using is provided by the world bank. The world bank does provide an API which we could query, but there is an R package available called wbstats which provides easy access to the data we want.
Information on using that wbstats package can be found here:

https://cran.r-project.org/web/packages/wbstats/vignettes/Using_the_wbstats_package.html

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#import our libraries
library(tidyverse) # tidyverse provides a beautiful set of tools for data science with consistent syntax and behaviour
library(wbstats) #wbstats provides access to the world bank data sets
library(knitr) #provides a set of tools to customize the processing of Rmarkdown including kable - a very simple table generator
library(kableExtra) # provides a set of tools to extend kable -
```

## Loading


```{r}
countries <- wb_cachelist$countries %>% as.tbl()
countries <- countries %>% 
  select(-c(capital,iso2c,region_iso2c,regionID,adminID,admin_iso2c,incomeID,income_iso2c,lendingID,lending_iso2c,lending,admin)) %>% 
  mutate(long = as.numeric(long), lat = as.numeric(lat))
countries %>% 
  head() %>% 
  kable()
```


```{r}
irrigation <- wb(indicator = "AG.LND.IRIG.AG.ZS") %>% #download the specified indicator from the World Bank API
  as.tbl() %>% #convert the dataframe to a tibble for better console printing in interactive analysis (limits default number of rows printed to a reasonable number and prints var type of each column beneath row name)
  mutate(date = as.numeric(date)) #convert the date column to a numeric value

irrigation %>% 
  head()%>% #return the first 6 rows of the irrigation tibble as a new tibble
  kable() #output the new tibble as an html friendly table
```



```{r}
tractors <- wb(indicator = "AG.LND.TRAC.ZS") %>% 
  as.tbl() %>% 
  mutate(date = as.numeric(date))
tractors %>% head() %>% kable()
```


```{r}
fertilizer <- wb(indicator = "AG.CON.FERT.ZS") %>% 
  as.tbl()%>% 
  mutate(date = as.numeric(date))
fertilizer  %>% head() %>% kable()
```


```{r}
cereal_yield <- wb(indicator = "AG.YLD.CREL.KG") %>% 
  as.tbl() %>% 
  mutate(date = as.numeric(date)) 
cereal_yield %>% head() %>% kable()
```


Some notes on what we can see so far:

 * We essentially have a number of data tables with the format [Country,Year,Indicator,Value] with some of these repeated in different formats. Needs reshaping.
 * We don't really have a very good name column at the moment. Need to make.
 * The country column for several of these datasets has a value "Arab World". As far as I am aware, that's not a country. Needs cleaning.

 
## Reshaping
 
```{r}
irrigation_red <- irrigation %>% 
  select(iso3c, country, date, value) %>% 
  rename(irrigation_percent_of_land = value)

irrigation_red
```

```{r}
tractors_red <- tractors %>% 
  select(iso3c, country, date, value) %>% 
  rename(tractors_per_100sqkm = value)

tractors_red
```

```{r}
fertilizer_red <- fertilizer %>% 
  select(iso3c, country, date, value) %>% 
  rename(fertilizer_kg_per_hectare = value)
fertilizer_red
```

```{r}
cereal_yield_red <- cereal_yield %>% 
  select(iso3c, country, date, value) %>% 
  rename(cereal_yield_kg_per_heactare = value)
cereal_yield_red
```

## Removing non countries

The easiest way to remove the non-countries from the data set is to join onto the countries dataframe.









```{r}
df <- countries %>% left_join(irrigation_red %>% select(-country),by="iso3c") %>% 
    left_join(tractors_red %>% select(-country),by=c("iso3c","date")) %>% 
    left_join(fertilizer_red %>% select(-country),by=c("iso3c","date")) %>% 
    left_join(cereal_yield_red %>% select(-country),by=c("iso3c","date")) %>% head() %>% kable()

```



```{r}

```

